---
title: "An Analysis of Water Quality in the Ogallala Aquifer"
author: "Justin Howard, Gavin Hudgeons, Eric Balke"
date: "June 6, 2019"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir ='C:/Users/howar/Documents/r__working_directory/Water_Project')
```

```{r, include=F, warning=F}
library(tidyverse)
library(stringr)
library(dplyr)
library(lubridate)
```

```{r well descriptions table, include=F}
wellmain<- read.table("C:/Users/howar/Documents/Stats.2/Project 1/WellMain.txt", sep="|", header=T,
                      stringsAsFactors = T, strip.white = T, quote="", comment.char = "", fill = T)
```
```{r loading water quality data, include=F}
qualitymajor<- read.table("C:/Users/howar/Documents/Stats.2/Project 1/WaterQualityMajor.txt", sep="|",
                          header=T,stringsAsFactors = T, strip.white = T, quote="", comment.char = "", 
                          fill = T)

qualitycombination<- read.table("C:/Users/howar/Documents/Stats.2/Project 1/WaterQualityCombination.txt",
                                sep="|", header=T, stringsAsFactors = T, strip.white = T, quote="",
                                comment.char = "", fill = T)

#loading levels data
levelsmajor<- read.table("C:/Users/howar/Documents/Stats.2/Project 1/WaterLevelsMajor.txt", sep="|",
                          header=T,stringsAsFactors = T, strip.white = T, quote="", comment.char = "", 
                          fill = T)
levelscombo<- read.table("C:/Users/howar/Documents/Stats.2/Project 1/WaterLevelsCombination.txt", sep="|",
                          header=T, stringsAsFactors = T, strip.white = T, quote="", comment.char = "", 
                          fill = T)
```
## Introduction
The Ogallala Aquifer underlies approximately 175,000 square miles and provides water for commercial, residential and agricultural customers across the High Plains. It is the primary water source for portions of Texas, New Mexico, Oklahoma, Kansas, Nebraska, Colorado, Wyoming and parts of South Dakota. According the a US Geological Survey report issued on June 16, 2017, it is also in decline.

This project constsists of a linear regression analysis of water quality (total dissolved solids (mg/L)) and a two-way analysis of variance of water level (ft) trends in the Panhandle Region of Texas, from 1925 to 2016. We hope to identify linear trends in water quality, with the goal of identifying factors that will best predict water quality in the coming years.

## Data Description
Our data comes from the Texas Groundwater database: http://www.twdb.texas.gov/groundwater/data/gwdbrpt.asp
We obtained observational data on each well in the state, including water quality samples and water level measurements.

<center>
#### Problem 1:
#### What factors can be used to predict water quality as measured by Total Dissolved Solids (mg/L)?
</center>
#### Approach:
We are choosing to limit the scope of this analysis to the Ogallala Aquifer, on wells located in Lubbock County, located in the Panhandle region of Texas. This aquifer was picked because its level of recharge is very slow, limiting the effect of rainfall on the analysis. Lubbock County was selected because it contains a high population density for the area. 

We limited the selection of observations to those with Well Depth readings, gps coordinates and added a variable to account for the mean water levels measured across all wells in Lubbock County for each year we had measurements of total dissolved solids.
```{r isolating data on the Ogallala Aquifer on wells in Lubbock County, include=F}
# water quality -- the qualityminor and qualityother tables had no ogallala
ogaqualitymajor<- qualitymajor %>%
  filter(str_detect(Aquifer, "Ogallala")) %>%
  filter(str_detect(County, "Lubbock"))
ogaqualitycombo<- qualitycombination %>%
  filter(str_detect(Aquifer, "Ogallala")) %>%
  filter(str_detect(County, "Lubbock"))
ogaqualityAll<- list(ogaqualitycombo, ogaqualitymajor)
ogaqualityAll<- do.call("rbind", ogaqualityAll)

# water levels
ogalevelsmajor<- levelsmajor %>%
  filter(str_detect(Aquifer, "Ogallala")) %>%
  filter(str_detect(County, "Lubbock"))
ogalevelscombo<- levelscombo %>%
  filter(str_detect(Aquifer, "Ogallala")) %>%
  filter(str_detect(County, "Lubbock"))
ogalevelsAll<- list(ogalevelscombo, ogalevelsmajor)
ogalevelsAll<- do.call("rbind", ogalevelsAll)

# getting aggregate values of water elevation by year
levelsbyyear<- aggregate(WaterElevation ~ MeasurementYear, ogalevelsAll, mean)
names(levelsbyyear)<- c("SampleYear", "MeanWaterLvl")

# combining tables to get more variables to select from
final<- merge(ogaqualityAll, wellmain, by = "StateWellNumber")
# filtering based on the response variable
finalTDS<- final %>%
  filter(str_detect(ParameterDescription, "TOTAL DISSOLVED SOLIDS"))
# filtering out wells without welldepth values
TDSDepth<- finalTDS %>%
  drop_na(WellDepth)

# aggregate to get the mean TDS per year
TDSbyyear<- aggregate(ParameterValue ~ SampleYear, finalTDS, mean)

# merging TDSbyyear with TDSDepth on SampleYear
aggTDS<- merge(TDSDepth, TDSbyyear, by = c("SampleYear"))

# merging TDS table with aggreagate water levels
TDS_Waterlvl<- merge(TDSDepth, levelsbyyear, by = "SampleYear")

```

```{r panel definition for plotting the response, include=F, warning=F}
attach(TDS_Waterlvl)
panel.cor <- function(x, y){
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- round(cor(x, y), digits=2)
    txt <- paste0("R = ", r)
    text(0.5, 0.5, txt)
}
```
## Exploratory Data Analysis
```{r plotting the repsonse, echo=F, warning=F}
pairs(~ ParameterValue + SampleYear + LandSurfaceElevation + LatitudeDD + LongitudeDD + WellDepth
      + MeanWaterLvl, 
      lower.panel = panel.cor,
      data= TDS_Waterlvl)
```
## Objective 1: Highly Interpretable Regression Analysis
#### Methodology:
Correlations were added to the plot to help determine covariance between predictors. Calculations are based on Pearson's R and high values will help determine placement in the model.

#### Observations:
The correlations beween GPS coordinates and land elevation are high. A previous study of water quality found signficant evidence that that wells in the northern portion of the Ogallala Aquifer had overall lower TDS levels than those in the southern portion. To see if this holds true for Lubbock County as well, we will keep latitude and longitude coordinates over land elevation. LandSurfaceElevation may need to be removed from the model. 
MeanWaterlvl is highly corrleated with Year, so Year will be removed from the model, but investigated for trends over time.
Also of note: A signficant outlier appears to to be in a sample taken in year 1980.
```{r boxplot ParameterValue, echo=F}
library(car)
  Boxplot(TDS_Waterlvl$ParameterValue, 
          main="Boxplot of ParameterValue",
          ylab = "Total Dissolved Solids (mg/L)",
          id.method = 'y')
## performing log transformation on response
    Boxplot(log(TDS_Waterlvl$ParameterValue), 
          main="Boxplot of log(ParameterValue)",
          ylab = "Total Dissolved Solids",
          id.method = 'y')
```

#### Observations:
The transformation has improved the distance between the outliers and the mean value. These observations will be removed to judge their effect on the model.

```{r removing outliers, include=F}
outliers<- c(215, 238, 202, 190, 189, 212, 180, 193, 204, 207)
noOutliers<- TDS_Waterlvl[-outliers,]
```
```{r outliers removed, echo=F}
    Boxplot(log(noOutliers$ParameterValue), 
          main="Boxplot of log(ParameterValue)",
          ylab = "Total Dissolved Solids",
          id.method = 'y')
```

#### Observations:
Standard deviation has improved significantly, the outliers will be removed from the training data.
```{r scatterplots for categoricals, echo=F}
pairs(~ParameterValue + ReportingAgency + Owner + Pump + PowerType + WellUse + WellType + AnalyzedLab,
      #lower.panel is a Pearson's R correlation 
      lower.panel = panel.cor,
      data = noOutliers)
```

#### Observations:
Aquifer and Classification appear to be strongly correlated, as do SampledAquifer and Classification. 

WellUse will stay in the model to control for the type of well being sampled from, but Pump and PowerType will be removed due to strong corrleations with WellType and WellUse. 

ReportingAgency, AnalyzedLab and Owner may cause the model to be overfit given that they are not helpful in predicting water quality, but we may want to account for peculiarities in reporting statistics. 

WellType will be removed as there only appears to be one type of well in the data.

Before we proceed, we will look at scatterplots of a log transformed repsonse to see if outliers are dealt with.
```{r plots of log transformed ParameterValue, warning=F, echo=F}
pairs(~log(ParameterValue) + LatitudeDD + LongitudeDD + WellDepth + MeanWaterLvl, 
      lower.panel = panel.cor,
      main = "Log(TDS) v. Continuous Variables",
      data = noOutliers)
pairs(~log(ParameterValue) + ReportingAgency + Owner + WellUse + AnalyzedLab + SampleYear 
      + LatitudeDD + LongitudeDD,
      lower.panel = panel.cor,
      main = "Log(TDS) v. Categorical Variables",
      data = noOutliers)
```

#### Observations:
A small correlation exists between Owner and WellUse, but this may be insignificant. Models with and without Owner will be compared.
Strong correlations exist between SampleYear, AnalyzedLab and ReportingAgency. AnalyzedLab and ReportingAgency will be removed.

```{r linear model with outliers}
#model with outliers
model1<- lm(log(ParameterValue)~ + LatitudeDD + LongitudeDD + Owner + WellUse + WellDepth
            + MeanWaterLvl, data = TDS_Waterlvl)
plot(model1$fitted.values, model1$residuals, main = "Model1, Outliers Included", xlab="Fitted Values", ylab="Residuals")
```

#### Observations:

The residuals plot shows clear evidence that the leverage of the outliers is significant. This confirms the previous decision to remove the outliers from the model.
```{r linear model no outliers, Owners removed}
# Outliers removed
model2<- lm(log(ParameterValue)~ + LatitudeDD + LongitudeDD + WellUse + WellDepth
            + MeanWaterLvl, data = noOutliers)
plot(model2$fitted.values, model2$residuals, main = "Model2, Outliers & Owners Removed", xlab="Fitted Values", ylab="Residuals")
```

#### Observations:
The residuals are much more visible. There is some evidence of clustering and non-constant variance, but the plot has improved significantly.

A closer investigation of the continuous variables will be helpful.
```{r searching for non-constant variance, echo=F}
par(mfrow=c(1,2))
plot(log(ParameterValue) ~ LatitudeDD, data = noOutliers)
plot(log(ParameterValue) ~ LongitudeDD, data = noOutliers)
par(mfrow=c(1,2))
plot(log(ParameterValue) ~ SampleYear, data = noOutliers)
plot(log(ParameterValue) ~ MeanWaterLvl, data = noOutliers)
```

#### The scatterplots indicate 4 important relationships:
1. As latitude increases, TDS, decreases (the aforementioned north-south degredation in water quality exists on the county level)
2. As longitude increases, TDS decreases (an east-west degredation in water quality also exists.)
3. Serial corrleation is evident. Since 1940, TDS values trend upward. The water quality of these wells is decreasing.
4. As mean water levels decrease, so does water quality as measured by TDS.

Year and MeanWaterLvl are too highly correlated. Year will be removed, because water levels are more highly corrleated with TDS levels than Year.

## Preliminary Linear Model Using Custom Feature Selection

```{r model1 summary, include=F}
summary(model1)
```

#### Observations from Preliminary Model1 summary:
The adjusted R-squared for model 1 is .72. Including well owners will overfit the model and are not great predictors of water quality. They will be taken out of the model.

```{r model 2 summary, include = F}
summary(model2)
```

#### Observations from Preliminary Model2 summary:
Model 2 has a much worse R squared value (.4273). This is an acceptable sacrifice as Owners are not valuable to the analysis of water quality.

```{r training and test set split, include=F, echo=T}
set.seed(1234)
#creating an 80:20 train:test split
sample_size <- floor(0.80 * nrow(noOutliers))
index<- sample(seq_len(nrow(noOutliers)), size = sample_size)
outDF<- TDS_Waterlvl[outliers,]
train<- noOutliers[index,]
test<- noOutliers[-index,]
```

#### Observations:
Before returning the outliers to the test set, an examination revealed that 9 out of 10 outliers were sampled in the year 1980. 8 samples were from well 2326306, belonging to a the Lubbock Children's Home. 2 samples were from the Benton Estate. All samples had a variant of the same note, "Sample collected from well...not filtered or preserved." These outliers appear to be caused by a malfunction of sorts and will not be reintroduced to the model.

For the automatic feature selection, we will put LandElevation back into the model to allow the algorithms more flexibility.

## Automatic Feature Selection
```{r creating predictor and response variables}

#Predictors saved as object
predictors<- model.matrix(~ + LatitudeDD + LongitudeDD + WellUse + WellDepth + LandSurfaceElevation 
                          + MeanWaterLvl + WellType, train) [,-1]

response<- log(train$ParameterValue)
dim(predictors)
dim(response)
```

## Ridge Selection
```{r ridge selection, warning=F}
library(glmnet)
cv.ridge <- cv.glmnet(predictors, response, alpha = 0)
cv.ridge$lambda.min
model.ridge <- glmnet(predictors, response, alpha = 0, lambda = cv.ridge$lambda.min)
plot(cv.ridge)
coef(model.ridge)
```

#### Observations:
The ridge method selects 14 features. The top 3 predictors of water quality were:
1. longitude (-2.236)
2. WellUse Recreation (.506)
3. latitude (.606)

```{r ridge model test, warning=F, eval=F}
library(tidyverse)
library(caret)
test.ridge <- model.matrix(~ + LatitudeDD + LongitudeDD + WellUse +WellDepth + LandSurfaceElevation 
                          + MeanWaterLvl + WellType, test) [,-1]
predictions.ridge <- model.ridge %>%
  predict(test.ridge) %>% 
  as.vector()
data.frame(
  RMSE.r = RMSE(predictions.ridge, log(test$ParameterValue)),
  Rsquare.r = R2(predictions.ridge, log(test$ParameterValue)))
```

#### Observations:
Ridge selection out-performs the custom model selection from Model2. A notable deviation is that the algorithm found LandSurfaceElevation, Latitude and Longitude as statistically significant even though they are correlated with each other.

## LASSO Selection
```{r LASSO selection and testing, eval=F}
cv.l <- cv.glmnet(predictors, response, alpha = 1)
cv.l$lambda.min
plot(cv.l)
model.lasso <- glmnet(predictors, response, alpha = 1, lambda = cv.l$lambda.min)
lasso.indicators<-coef(model.lasso)
lasso.indicators
test.lasso <- model.matrix(~ + LatitudeDD + LongitudeDD + WellUse +WellDepth + LandSurfaceElevation 
                          + MeanWaterLvl + WellType, test)[,-1]
predictions.lasso <- model.lasso %>%
  predict(test.lasso) %>% 
  as.vector()
data.frame(
  RMSE.l = RMSE(predictions.lasso, test$ParameterValue),
  Rsquare.l = R2(predictions.lasso, test$ParameterValue))
```

#### Observations:
The LASSO method found 13 features significant. The top 3 predictors of water quality were:
1. longitude (-3.547)
2. latitiude (1.509)
3. WellUse Recreation (.4523)
```{r Elastic Net selection and testing, eval=F}
model.net <- train(log(ParameterValue) ~ + LatitudeDD + LongitudeDD + WellUse +WellDepth 
                   + LandSurfaceElevation + MeanWaterLvl + WellType, 
                   data = train, 
                   method = "glmnet",
                   trControl = trainControl("cv", number = 10), 
                   tuneLength = 10)
model.net$bestTune
  
coef(model.net$finalModel, model.net$bestTune$lambda)
test.net <- model.matrix(log(ParameterValue) ~ + LatitudeDD + LongitudeDD + WellUse + WellDepth 
                         + LandSurfaceElevation + MeanWaterLvl + WellType, data = test)[,-1]
plot(coef(model.net$finalModel))
#predictions.net <- model.net %>% 
#  predict(test.net)
# data.frame(
# RMSE.net = RMSE(predictions.net, test$ParameterValue),
# Rsquare.net = R2(predictions.net, test$ParameterValue))
```

####Observations:
The Elastic Net method found 13 features significant. The top 3 predictors of water quality were:
1. longitude (-3.504)
2. latitiude (1.475)
3. WellUse Recreation (.4526)
Unfortunately, time does not permit the troubleshooting of the elastic net RMSE and Rsquared, so this model will not be selected. 

## Parameter Interpretations
The ridge model provided the best combined R-squared and RMSE values and the predictors of water quality from this model are interpreted as follows, with the coefficients in the middle and the interpretations (interpreted as 1-exp(value)):

For a 1 unit increase in:              The median total dissolved solids is expected to:

LatitudeDD                   1.834      increase 83.4%
LongitudeDD                  0.752      decrease 89.3%
LandSurfaceElevation         0.996      decrease 4%
MeanWaterLvl                 0.996      decrease 4%
WellDepth                    0.998      decrease 2%

### Water Sources with elevated total dissolved solids    

Source:                      Change in median total dissolved solids:

WellUseDomestic              0.969    decrease 3.1%
WellUseIndustrial            0.752    decrease 24.8%
WellUseInstitution           0.857    decrease 14.3%
WellUseIrrigation            1.040    increase 4.0%
WellUsePlugged or Destroyed  1.089    increase 8.9%
WellUsePower                 1.303    increase 30.3%
WellUsePublic Supply         0.957    decrease 4.3%
WellUseRecreation            1.658    increase 65.8%
WellUseUnused                0.962    decrease 3.8%



## Conclusions:

We identified the top 14 features that predict the median measurements of total dissolved solids in Lubbock County. The top predictor of total dissolved solids in Lubbock County is latitude. Every unit increase in latitude in Lubbock County is associated with a 83.4% increase in the median TDS value. Each unit increase in longitude is associated with an 89.3% decrease in median TDS values. It is important to note that increases in total dissolved solids are indicators of reduced water quality. Residents of south western Lubbock County can expect much worse water than in north eastern Lubbock county, if they drill their own water well.

Residents of Lubbock County can get better water quality, if they draw their water from a well correlated with higher quality water. Industrial wells are associated with a 24.8% decrease in TDS, while public water wells are associated with a 4.3% decrease in median TDS values. The worst water quality is obtained from wells used for recreation, (possibly to supply swimming pools), with a 65.8% increase in median TDS and wells used for power, with a 30.3% increase in median TDS. 


#### Important Relationships:
1.	As latitude increases, TDS, decreases (the aforementioned north-south degredation in water quality exists      on the county level)
2.	As longitude increases, TDS decreases (an east-west degredation in water quality also exists.)
3.	Serial corrleation is evident. Since 1940, TDS values trend upward. The water quality of these wells is        decreasing with time.
4.	As mean water levels decrease, so does water quality as measured by TDS.

## Objective 2: Highly Predictive Regression Analysis
### Regression Forest with Grid tuning
#### Random Forest (method = 'rf')
Type: Classification, Regression
Tuning parameters:
mtry (#Randomly Selected Predictors)
```{r rf2, cache=FALSE}
# Random forest regression with Grid tuning across other possible values of hyperparameters
tg2 <- data.frame(mtry = seq(1, 25, by = 2))
rf2 <- caret::train(log(ParameterValue) ~ LatitudeDD + LongitudeDD + WellUse + WellDepth + LandSurfaceElevation + MeanWaterLvl + WellType, data = train, method = "rf", tuneGrid = tg2)
summary(rf2)
get_best_result(rf2)
rf2fin <- rf2$finalModel
```

Plot training results and calculate test data fit statistics
```{r rf2_plots and testing}
ggplot(rf2) + theme_bw() + ggtitle(wrapper("Random Forest Regression with Grid Search for finding optimal hyperparameters in caret::train method=rf", width = 60))

rf2predvalues_train<-predict(rf2, train)
plot(log(train$ParameterValue),rf2fin$residuals)

plot(rf2$finalModel)
rf2values_train <-data.frame(obs = log(train$ParameterValue), pred=rf2predvalues_train)
defaultSummary(rf2values_train)

# get stats on test set
rf2predValues_test <- predict(rf2, test)
rf2values_test <-data.frame(obs = log(test$ParameterValue), pred=rf2predValues_test)
defaultSummary(rf2values_test)
```

### Tree Models from Genetic Algorithms
#### Tree Models from Genetic Algorithms (method = 'evtree')
For classification and regression using package evtree with tuning parameters:
Complexity Parameter (alpha, numeric)
```{r evtree, cache=FALSE}
cctrl1 = trainControl(method="repeatedcv", number=5, repeats=3)
tune_grid_evtree <- expand.grid(alpha = seq(0.2, 0.45, by = 0.05))
evtree <- train(log(ParameterValue) ~ LatitudeDD + LongitudeDD + WellUse + WellDepth + LandSurfaceElevation + MeanWaterLvl + WellType, 
                      data = train, 
                      method = "evtree", 
                      trControl = cctrl1, 
                      tuneGrid = tune_grid_evtree)

evtree
summary(evtree)
get_best_result(evtree)
evtreefin <- evtree$finalModel
```

Plot training results and calculate test data fit statistics
```{r evtree plots and testing}
ggplot(evtree) + theme_bw() + ggtitle(wrapper("Tree model with Genetic Algorithms with Grid Search for finding optimal hyperparameters in caret::train method=evTree", width = 60))

evtreepredvalues_train<-predict(evtree, train)
plot(log(train$ParameterValue),evtreefin$residuals)
plot(evtreefin)

evtreevalues_train <-data.frame(obs = log(train$ParameterValue), pred=evtreepredvalues_train)
defaultSummary(evtreevalues_train)

# get stats on test set
evtreepredValues_test <- predict(evtree, test)
evtreevalues_test <-data.frame(obs = log(test$ParameterValue), pred=evtreepredValues_test)
defaultSummary(evtreevalues_test)
```

### Try stepwise regression with AIC on selected variables training dataset
#### Generalized Linear Model with Stepwise Feature Selection (method = 'glmStepAIC')
Type: Regression, Classification
No tuning parameters for this model

```{r EB_SW_AIC, cache=FALSE}
# train model
step_lm <- caret::train(log(ParameterValue) ~ LatitudeDD + LongitudeDD + WellUse + WellDepth + LandSurfaceElevation + MeanWaterLvl + WellType, 
                        data = train, 
                        method = "lmStepAIC", 
                        trControl = cctrl1, 
                        trace = FALSE)

summary(step_lm)
get_best_result(step_lm)
step_lm_fin <- step_lm$finalModel
step_lm_fin
```

Plot training results and calculate test data fit statistics
```{r SW AIC plots and testing}
step_lmres<-resid(step_lm)
step_lmpredvalues_train<-predict(step_lm, train)
plot(log(train$ParameterValue),step_lm_fin$residuals)
abline(0,0)
plot(step_lm_fin$fitted.values,step_lm_fin$residuals)

step_lmvalues_train <-data.frame(obs = log(train$ParameterValue), pred=step_lmpredvalues_train)
defaultSummary(step_lmvalues_train)

# get stats on test set
step_lmpredValues_test <- predict(step_lm, test)
step_lmvalues_test <-data.frame(obs = log(test$ParameterValue), pred=step_lmpredValues_test)
defaultSummary(step_lmvalues_test)
```

## Code Appendix
```{r ref.label=knitr::all_labels(), echo = T, eval = F}
